
---
title: "RMSTdesign Analysis Report"
output: pdf_document
params:
  inputs: NA
  results: NA
  log: NA
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(survival)
library(survminer)
library(kableExtra)
```

# Analysis Configuration

This report summarizes the power or sample size analysis performed using the RMSTdesign application.

```{r input-parameters}
inputs_df <- data.frame(
  Parameter = names(params$inputs),
  Value = unlist(lapply(params$inputs, function(x) paste(x, collapse = ", ")))
)

kbl(inputs_df, 
    booktabs = TRUE, 
    caption = "Analysis Input Parameters",
    col.names = c("Parameter", "Value")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

# Survival Analysis of Pilot Data

The following plot and table are based on the survival analysis of the uploaded pilot data, using the log-rank test.

```{r km-plot, eval = !is.null(params$results$analysis_data_for_plot)}
plot_data <- params$results$analysis_data_for_plot
fit <- survfit(Surv(time, status) ~ arm, data = plot_data)

p <- ggsurvplot(
  fit,
  data = plot_data,
  palette = c("#007BFF", "#D9534F"),
  legend.title = params$inputs$arm_var,
  xlab = paste("Time (in units of '", params$inputs$time_var, "')"),
  ylab = "Survival Probability",
  ggtheme = theme_light()
)

p$plot
```

```{r logrank-summary, eval = !is.null(params$results$logrank_summary)}
kbl(params$results$logrank_summary, 
    booktabs = TRUE, 
    caption = "Log-Rank Test Results on Pilot Data") %>%
  kable_styling(latex_options = "hold_position")
```

# Power and Sample Size Analysis

```{r power-curve, eval = !is.null(params$results$results_plot)}
params$results$results_plot
```

```{r results-table, eval = !is.null(params$results$results_data)}
kbl(params$results$results_data, 
    booktabs = TRUE, 
    caption = "Power/Sample Size Calculation Results") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r effect-size, eval = !is.null(params$results$results_summary)}
kbl(params$results$results_summary, 
    booktabs = TRUE, 
    caption = "Effect Size Summary from Pilot Data") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

# Console Log

The following is the complete console log generated during the analysis.

```{r console-log, results='asis', eval = !is.null(params$log)}
cat(params$log)
```
